---
title: Health and economic consequences from significant meteorological events in
  USA
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache = TRUE)
```


#Synopsis
In this document we will asses the effect of weather related events either those which are harmful to the people (cause of injuries and fatalties) or those which carry a cost due to property and crop damage. We analyze the data from the [ U.S. National Oceanic and Atmospheric Administration's (NOAA) storm database](http://www.nws.noaa.gov/directives/) and explore such effects across USA. We therefore conclude:

* Accros the United States the most harmful weather phenomena is the **Tornado** both in number of injuries and casualties.
* The major impact in property and crop damage from these extraordinary meteorological events is due to **floods** (highest *property damage costs*) and **drought** (highest *crop damage costs*).  

Thus this analysis of the NOAA database (link to the analyzed database can be found [here](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2)) sheds light on both in healthy and economic repercussion of this weather events.

#Data Processing
In this section we will describe the command executed in R to download and read the file into R and later we will describe how the data obtained is tidied and processed to be suitable to analysis (Which we will see in the results section)

##Obtaining and loading the data
```{r download, cache=TRUE, results='hide'}
# Downloading the file and documentation
fileURL <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
docURL <- "https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf"

if(!file.exists("./repdata%2Fdata%2FStormData.csv.bz2")) {
        download.file(fileURL, destfile = "repdata%2Fdata%2FStormData.csv.bz2")
}

if(!file.exists("./repdata%2Fpeer2_doc%2Fpd01016005curr.pdf")) {
        download.file(docURL, destfile = "./stormdata.pdf")
}

#Loading data into R
stormdata <- read.csv("repdata%2Fdata%2FStormData.csv.bz2")

#Loading neccesary packages
library(dplyr)
library(RColorBrewer)

```

##Processing the data
Since the names of variables are not in an easy way to work with, we will modificed them:
* The names of the variables seem to have a strange format:
```{r names}
names(stormdata)
```
* We will process these names in order to be easy to work with (NOTE:_These steps are a personal preference, the data can be analized without these steps_):
```{r tidynames}
cnames <- names(stormdata)
cnames <- gsub("_", ".", cnames)
cnames <- tolower(cnames)
cnames[1] <- "state"
cnames[cnames == "longitude."] <- "longitude.e"
cnames #Lest see how it looks now
names(stormdata) <- cnames
```
* Now we will substract the variables we are interested in:
```{r variables}
data <- stormdata[, c("bgn.date", "evtype", "fatalities", "injuries", "propdmg", "propdmgexp", "cropdmg", "cropdmgexp")]
```
* We will prepare the data to analyze and plot the results regarding to harmful effects on human populations: 
```{r harmfultidy}
#Clean the data out of unnecessary rows
harmful <- data[(!is.na(data$injuries) & data$injuries > 0) | (!is.na(data$fatalities) & data$fatalities > 0), c("evtype", "injuries", "fatalities")]
```
* And now we will procedure to process the data in order to analyze the ecomony impact derived from weather events:
1. We need to merge the variables propdmg and propdmgexp. We first take a look to propdmgexp:
```{r prop}
prop <- data[(!is.na(data$propdmg) & data$propdmg > 0), c("evtype", "propdmg", "propdmgexp")]
levels(prop$propdmgexp)
```
2. And now the variables cropdmgexp:
```{r crop}
crop <- data[(!is.na(data$cropdmg) & data$cropdmg > 0), c("evtype", "cropdmg", "cropdmgexp")]
levels(crop$cropdmgexp)
```

* We will assume that the factors "K" (and k), "B" (and b), "M" (and m) and "H" (and h) stand for Thoundsan, Billions, Millions and Hundreds respectively (the rest will be associated to 1 as a value), which in combination with the values stored in the prop/cropdmg variable will give us the absulte values:
```{r combining}
# We will combine first the variables propdmg and propdmgexp
prop$propdmgexp <- as.character(prop$propdmgexp)
prop[prop$propdmgexp == "" |
             prop$propdmgexp == "-" | 
             prop$propdmgexp == "?" |
             prop$propdmgexp == "+" |
             prop$propdmgexp == "0" |
             prop$propdmgexp == "1" |
             prop$propdmgexp == "2" |
             prop$propdmgexp == "3" |
             prop$propdmgexp == "4" |
             prop$propdmgexp == "5" |
             prop$propdmgexp == "6" |
             prop$propdmgexp == "7" |
             prop$propdmgexp == "8",
     c("propdmgexp")
             ] <- "1"
prop[prop$propdmgexp == "H" | prop$propdmgexp == "h", c("propdmgexp")] <- 10^2
prop[prop$propdmgexp == "M" | prop$propdmgexp == "m", c("propdmgexp")] <- 10^6
prop[prop$propdmgexp == "B" , c("propdmgexp")] <- 10^9
prop[prop$propdmgexp == "K" , c("propdmgexp")] <- 10^3
prop$propdmgexp <- as.numeric(prop$propdmgexp)
prop <- mutate(prop, totaldmg = propdmg * propdmgexp) %>% select(evtype, totaldmg)

# And now the same with cropdmg and cropdmgexp
crop$cropdmgexp <- as.character(crop$cropdmgexp)
crop[crop$cropdmgexp == "" |
             crop$cropdmgexp == "?" |
             crop$cropdmgexp == "0" |
             crop$cropdmgexp == "2",
     c("cropdmgexp")] <- "1"
crop[crop$cropdmgexp == "B", c("cropdmgexp")] <- 10^9
crop[crop$cropdmgexp == "K" | crop$cropdmgexp == "k", c("cropdmgexp")] <- 10^3
crop[crop$cropdmgexp == "M" | crop$cropdmgexp == "m", c("cropdmgexp")] <- 10^6
crop$cropdmgexp <- as.numeric(crop$cropdmgexp)
crop <- mutate(crop, totaldmg = cropdmg * cropdmgexp) %>% select(evtype, totaldmg)
```

##Results
This section compromissed two basic question detailed below.

1. **Across the United States, which types of events (as indicated in the _evtype_ variable) are most harmful with respect to population health?**

We first look at the data we have processed to get an idea how it looks:
```{r harmful}
#Arranging the evtype data for injuries and fatalities
harmful <- group_by(harmful, evtype)
harmfuli <- summarise(harmful, injuries = sum(injuries)) %>% arrange(desc(injuries))
harmfulf <- summarise(harmful, fatalities = sum(fatalities)) %>% arrange(desc(fatalities))

#Make us an idea of how the data look before plotting
head(harmfuli, 10)
head(harmfulf, 10)
```

And since we want to see the most harmful ones we can select the first 10:
```{r top10}
#Selecting the 10 most harmful events in both data variables
harmfulimost <- harmfuli[1:10,]
harmfulfmost <- harmfulf[1:10,]
```

```{r harmful plotting, fig.height=7}
par(mfrow=c(2,1), mar = c(5, 5, 2, 0), oma = c(1, 0, 2, 1))

with(harmfulimost, barplot(injuries, main = "Injuries", font.main = 1, ylab = "Count", cex.names = .6, cex.axis = .6, names.arg = evtype, las = 2, col = brewer.pal(10, "Spectral")))

with(harmfulfmost, barplot(fatalities, main = "Fatalities", font.main = 1, cex.names = .6, ylab = "Count", cex.axis = .6, ylim = c(0, 80000), names.arg = evtype, las = 2, col = brewer.pal(10, "Spectral")))

mtext("Weather related harmful events", font.main = 1, cex = 1.5, outer = TRUE)
```

The answer seems clear: **Tornados** are the most deathly and dangerous weather events.

2. **Across the United States, which types of events have the greatest economic consequences?**
```{r mostdmg}
#Summarising the results
prop <- group_by(prop, evtype)
props <- summarize(prop, totaldmg = sum(totaldmg)) %>% arrange(desc(totaldmg))
crop <- group_by(crop, evtype)
crops <- summarize(crop, totaldmg = sum(totaldmg)) %>% arrange(desc(totaldmg))

#Now see how it looks
head(props, 10)
head(crops, 10)

#and we will select the most destructive ones
propmost <- props[1:10,]
cropmost <- crops[1:10,]
```

```{r plotmostdmg, fig.height=7}
par(mfrow=c(2,1), mar = c(6, 5, 2, 0), oma = c(1, 0, 2, 1))

with(propmost, barplot(totaldmg/1000000000, main = "Property damage", font.main = 1, ylab = "Quantity (Billions)", cex.names = .6, cex.axis = .6, names.arg = evtype, las = 2, col = brewer.pal(10, "BrBG")))

with(cropmost, barplot(totaldmg/1000000000, main = "Crop damage", font.main = 1, cex.names = .6, ylab = "Quantity (Billions)", cex.axis = .6, ylim = c(0, 140), names.arg = evtype, las = 2, col = brewer.pal(10, "BrBG")))

mtext("Events with greatest economic consequences", font.main = 1, cex = 1.5, outer = TRUE)
```

Here **Flood** is the main event causing property damage while **Drought** is responsable of the highest costs in crop damage
